---
title: "Mini Data Analysis 2"
Author: "Marc Wettengel"
Date: "October 16, 2021"
output: github_document 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include = FALSE}
# load necessary packages
library(datateachr)
library(tidyverse)
```

# Task 1: Process and Sumarize Data

## 1.1 - Research questions from Milestone 1

**1. How does tree density vary in each neighbourhood?**

**2. Are there patterns to the height class of evergreen vs. deciduous species in Vancouver?**

**3. How has the number of trees planted per year change over time, and are there differences between genus or neighborhood?**

**4. Is there a relationship, if there is how strong is the correlation, between tree diameter and height class?**

## 1.2 - Summarizing data for each research question 

  For each research question I will select one option from the summarizing and graphing lists below and conduct a simple analysis to answer each of them.

**Summarizing**

1. Compute the range, mean, and two other summary statistics of one numerical variable across the groups of one categorical variable from your data.
2. Compute the number of observations for at least one of your categorical variables. Do not use the function table()!
3. Create a categorical variable with 3 or more groups from an existing numerical variable. You can use this new variable in the other tasks! An example: age in years into “child, teen, adult, senior”.
4. Based on two categorical variables, calculate two summary statistics of your choosing.

**Graphing**

5. Create a graph out of summarized variables that has at least two geom layers.
6. Create a graph of your choosing, make one of the axes logarithmic, and format the axes labels so that they are “pretty” or easier to read.
7. Make a graph where it makes sense to customize the alpha transparency.
8. Create 3 histograms out of summarized variables, with each histogram having different sized bins. Pick the “best” one and explain why it is the best.



```{r}
#Want to begin by creating a column with just year (removing the month and day aspect to the date_planted variable)
vancouver_trees$year <- as.Date(vancouver_trees$date_planted, format = "%Y-%m-%d")
vancouver_trees$Year <- as.numeric(format(vancouver_trees$year, '%Y'))
head(vancouver_trees)
Trees <-subset(vancouver_trees, select = -c(year))
head(Trees)
```

### 1.2.1 Research Question 1

**Question:** How does tree density vary in each neighbourhood?

**Summarizing:** 2. Compute the number of observations for at least one of your categorical variables. Do not use the function table()! 

```{r}
#Because each tree is attached to a neighbourhood value (name of neighbourhood), we can just find the count of each neighbourhood in the tree data set and use that as the count of trees. BUT, we want to first remove all NA observations from the neighbourhood_name variable and then do the count!
Trees %>%
  filter(!is.na(neighbourhood_name)) %>%
  group_by(neighbourhood_name) %>%
  count(neighbourhood_name)
  
```

**Grpahing:** 6. Create a graph of your choosing, make one of the axes logarithmic, and format the axes labels so that they are “pretty” or easier to read.

```{r}
#Plotting the counts of trees in each neighbourhood. When finding the density of trees, we first need to know the count of how many observations are in each neighbourhood. Graphing helps us visualize the differences in neighbourhoods, but isn't conclusive (obviously) because we need to know the area of each neighbourhood. 
Trees %>%
  filter(!is.na(neighbourhood_name)) %>%
  ggplot() +
  geom_bar(aes(y = neighbourhood_name)) +
  ggtitle("Tree Count in Vancouver Neighbourhoods") +
  ylab("Neighbourhood Name") +
  xlab("Number of Trees")

#In Milestone 3, I will calculate the neighbourhood size.
  
```

### 1.2.2 Research Question 2

**Question:** Are there patterns to the height class of evergreens vs. deciduous species in Vancouver?

**Summarizing:** 1. Compute the range, mean, and two other summary statistics of one numerical variable across the groups of one categorical variable from your data. 

```{r}
# Due to the vast number of species of trees, I'm going to randomly select 5 genus's of deciduous and 5 genus's of evergreen species to compare.
deciduous <- Trees %>% filter(genus_name == c("QUERCUS", "ULMUS", "CORNUS", "KOELREUTERIA", "TILIA"))
  
evergreen <- Trees %>% filter(genus_name == c("PICEA", "PINUS", "PSEUDOTSUGA", "CEDRUS", "SEQUOIADENDRON"))
unique(deciduous$genus_name)
unique(evergreen$genus_name)

evded <- rbind(deciduous, evergreen)
head(evded)
```

```{r}
#Summary statistics for both the deciduous and evergreen species that were randomly selected. 
sumdec <- deciduous %>%
  filter(!is.na(height_range_id)) %>% #remove all NA values if there are any in the height ID variable
  summarise(
    count = n(),
    mean = mean(height_range_id),
    median = median(height_range_id),#categorical data so not a very good measure
    min = min(height_range_id),
    max = max(height_range_id),
    sd = sd(height_range_id)
    )

sumev <- evergreen %>%
  filter(!is.na(height_range_id)) %>% #remove all NA values if there are any in height ID variable
  summarise(
    count = n(),
    mean = mean(height_range_id),
    median = median(height_range_id),
    min = min(height_range_id),
    max = max(height_range_id),
    sd = sd(height_range_id)
    )

sumevded <- rbind(sumdec, sumev) #combine the summary stats for easy observations, differences or similarities between deciduous and evergreen genus's
head(sumevded)
```

**Grpahing:**

```{r}
#Want to create a new variable determining if the tree is an evergreen or a deciduous.
#evded is a combo of evergreen and deciduous
evded$type <- as.factor(ifelse(evded$genus_name == c("PINUS", "PICEA", "PSEUDOTSUGA", "CEDRUS", "SEQUOIADENDRON"), 'EVERGREEN','DECIDUOUS'))
head(evded)

#side by side box plots between deciduous and evergreen species for comparison. I included a geom_violin() function to see how similar the distributions of values are (patterns in density) between heights of evergreens and deciduous
evded %>%
  filter(!is.na(height_range_id)) %>%
  filter(!is.na(type)) %>%
  ggplot(aes(x = height_range_id, y = type)) +
  geom_violin()+
  geom_boxplot(width = 0.2) +
  ggtitle("Comparison of heights between Evergreen and Deciduous species") +
  xlab("Height Class ID")

```


### 1.2.3 Research Question 3

**Question:** How does the number of trees planted change over time, and are there differences between genus or neighborhood?

**Summarizing:** 2/3. Compute the number of observations for at least one of your categorical variables. Do not use the function table()! Additionally, create a categorical variable with 3 or more groups from an existing numerical variable.


```{r}
#creating three categories (bins) for the year the tree was planted
evdednew <-evded %>% 
  filter(!is.na(height_range_id)) %>% 
  filter(!is.na(Year))
max(evded$Year)
evdednew$age <- as.factor(ifelse(evdednew$Year < 1999, "90s",
                              ifelse(evdednew$Year < 2009, "00s",
                              ifelse(evdednew$Year < 2020, "10s")))) #chose 2020 to include 2019 value

#chose these three bins because the oldest tree was planted in 1989 and the goungest was 2019, so three bins would each capture one third of the total time in the dataset. since 1989 is the first and 2019 is the last, the bins seem like weird dates
evdednew %>% 
  group_by(type) %>%
  summarise(sum_of_old = sum(age == "90s"))

```

**Grpahing:** 5. Create a graph out of summarized variables that has at least two geom layers.

```{r}
#Using a subset data of evded for this plot since there are so many NA values in the evded dataset and is why we went from 4110 observations to just 1145. Because of the vast amount of data missing, I would not recommend any further analysis with this research question! (cannot tell if it was random or if there is a pattern of NA values)


ggplot(evdednew, aes(x = Year)) +
  geom_bar(color = "black", aes(fill = age)) +
  ggtitle("Number of trees planted each year in Vancouver Neighbourhoods") +
  ylab("Count")

  
```

### 1.2.4 Research Question 4

**Question:** Is there a relationship, if there is how strong is the correlation, between tree diameter and height class?

**Summarizing:** 1. Compute the range, mean, and two other summary statistics of one numerical variable across the groups of one categorical variable from your data.

```{r}
#This is not helpful but couldn't think of anything else to compute (that my skills allow at least)

#remove outliers
outliers <- boxplot.stats(evded$diameter)$out

evded %>% 
  filter(type == "EVERGREEN") %>%
  filter(!is.na(diameter)) %>% #remove all NA values if there are any
  filter(!diameter %in% outliers) %>% #No more outliers
  summarise(
    count = n(),
    mean = mean(diameter),
    median = median(diameter),
    min = min(diameter),
    max = max(diameter),
    sd = sd(diameter)
    )
head(evded)
```

**Grpahing:**

```{r}
#creating a scatterplot to visualize the relationship between tree Diameter and the height class ID. 
evded %>%
  filter(!is.na(diameter)) %>%
  filter(!is.na(height_range_id)) %>%
  filter(!is.na(type)) %>%
  filter(!diameter %in% outliers) %>% #Filter out the outliers!
  ggplot(aes(height_range_id, diameter, color = type)) +
  geom_point(alpha = 0.6) + 
  geom_smooth(method = "lm") + #adding in a linear regression line for decisuous and evergreen species
  ggtitle("Relationship between Height ID and Diameter of trees in Vancouver") +
  xlab("Height ID") +
  ylab("Diameter (cm?)") 

  
```

## 1.3 

Here are the 4 research question I had initially:

1. How does tree density vary in each neighbourhood?
2. Are there patterns to the height class of evergreens vs. deciduous species in Vancouver?
3. How does the number of trees planted change over time, and are there differences between genus or neighborhood?
4. Is there a relationship, if there is how strong is the correlation, between tree diameter and height class?

Upon completing this milestone, I can eliminate one of my research questions I had prior to this assignment. Question 3 can be removed from my scope because there are a lot of values in my subset of the **vancouver_trees** data set for the year variable that were missing. After removing the observations without a year value, the length of the data set was approximately one fourth of the original. With so many observations lost, I cannot determine if there was a random distribution of trees without dates or if there was some limitation specific to a certain category of trees or other variable present in the data set. 

Research question 1 is most interesting to me, however, the challenge will be finding the area for each neighbourhood in Vancouver. I have an idea on how to complete this (however not very accurate) but will be good a exercise to further my skills in R and tidy data. Questions 1 and 4 are yielding the most interesting results to me and both pose unique, but intriguing challenges in my analysis project. I am much closer to answering these questions, I just need to overcome an obstacle in each anaylsis to yield a complete result. 

# Task 2: Tidy your data

## 2.1 Is my data Tidy?

A reminder of the definition of tidy data:

 - Each row is an **observation**
 - Each column is a **variable**
 - Each cell is a **value**

```{r}
#before determining if the data set is tidy, I will use the head() command to visualize my data set. 
head(vancouver_trees)
```

```{r}
#There are more than 8 columns, so I will select the 8 that I will most likely use to answer my 2 research questions of interest.


vancouver_trees %>%
  select(genus_name,
         species_name,
         common_name,
         neighbourhood_name,
         height_range_id,
         diameter,
         date_planted,
         longitude,
         latitude,
         year
         ) %>%
  na.omit() -> Vancity_trees
head(Vancity_trees)
```

Now we can see if the dataset is tidy or not. We can answer the questions posed to us earlier:

**1. Is each row an observation?**

  It is very clear that each row of this data set is an observation as each row is a unique tree in the city of Vancouver, indicated by the unique tree ID number in the original dataset (I removed the variable because it wouldn't be useful in my analysis)
  
**2. Is each column a variable?**

  Each column represents a variable in this data set and each variable is independent (meaning it has it's own row) in the data set.
  
**3. Is each cell a value?**

  Each cell contains a value (depending on the varible it's either categorical or continious) for each observation present in the data set, so this requirement for tidy data is met!
  
## 2.2 Untidy and retidy **Vancity_trees**
 
```{r}
untidyvancity <- Vancity_trees %>%
  pivot_wider(names_from = neighbourhood_name,
                values_from = height_range_id)
glimpse(untidyvancity)
head(untidyvancity)
```

This data now looks untidy and most cell values don't exist anymore! This data set is not optimal for my analysis and some values I need aren't included anymore from the original **vancouver_trees** data set.

```{r}
#now we want to retidy the data set! 
tidyvancity <- untidyvancity %>%
  pivot_longer(cols = -c(genus_name:year),
               names_to = "neighbourhood_name",
               values_to = "height_range_id")
head(tidyvancity)
glimpse(tidyvancity)
#YAY
```

## 2.3 Choose 2 research questions

The two research questions I will continue my analysis on are:
  
  **1. How does tree density vary in each neighbourhood?**
  
  **4. Is there a relationship, if there is how strong is the correlation, between tree diameter and height class?**
  
I chose both of these questions partly because I had to remove question 3 from my scope of interest due to the elimination of many observations after tidying the data set. Beyond that, these two questions pose the most unique challenges to me, and have attributes I enjoy about them. I've only lived in Vancouver for a few months now and I noticed how each neighbourhood has a different distribution of trees and species and I want to see if those differences are also apparent in this data set. Question 4 I want to analyze more because there are a lot of other questions that can stem from a correlation and scatterplot of tree diameter vs. height and so I want to further dive into this question.

### Data set to use in Milestone 3
```{r}
#select all potential variables I may need to answer my questions in Milestone 3
vancouver_trees %>%
  select(genus_name,
         neighbourhood_name,
         height_range_id,
         diameter,
         longitude,
         latitude,
         ) %>%
  #Remove all NA values present 
  na.omit() -> vantreeM3
head(vantreeM3)
```

### Data for question 1

```{r}
# for question 1 I just want to select all variables I will need as I don't know the proper mutate() command to finish tidying my data set.
vantreeQ1 <- vantreeM3 %>%
  select(genus_name, #including genus name just in case!
         neighbourhood_name,
         longitude,
         latitude)

# Finding how many trees are present in each neighbourhood. Since every tree is attached to a neighbourgood I can just count each neighbourhood observation as a tree!
ntrees <- vantreeQ1 %>%
  group_by(neighbourhood_name) %>%
  count(neighbourhood_name)
#Won't combine with other data set due to them being different lengths, but us an organized wat to calculate counts!
head(vantreeQ1)
```

### Data for question 4

```{r}
#For question 4, want to remove diameter outliers since I noticed them earlier.
outliersM3 <- boxplot.stats(vantreeM3$diameter)$out

#also selecting the variables I will use in my analysis
vantreeQ4 <- vantreeM3 %>%
  select(height_range_id,
         diameter,
         genus_name #including just in case!
         ) %>% 
  filter(!diameter %in% outliersM3) #bye bye outliers
```

















































